---
title: "Automatic Variable Labeling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Automatic Variable Labeling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options(gtsummary.print_engine = "gt")
```

```{r setup}
#| eval: false
library(sumExtras)
library(gtsummary)
library(dplyr)

use_jama_theme()
```

```{r setup2}
#| echo: false
#| message: false
#| warning: false
library(sumExtras)
library(gtsummary)
library(dplyr)
library(gt)

use_jama_theme()
```

Raw variable names like `trt`, `marker`, and `grade` don't belong in a publication table. If you're building 20+ tables across an analysis, manually relabeling the same variables in every `tbl_summary()` call is a waste of time. `add_auto_labels()` lets you define labels once and apply them everywhere.

## Creating a Data Dictionary

A dictionary is a data frame with two columns: `Variable` (exact variable names, case-sensitive) and `Description` (the labels you want displayed).

```{r}
dictionary <- tibble::tribble(
  ~Variable,    ~Description,
  "trt",        "Chemotherapy Treatment",
  "age",        "Age at Enrollment (years)",
  "marker",     "Marker Level (ng/mL)",
  "stage",      "T Stage",
  "grade",      "Tumor Grade",
  "response",   "Tumor Response",
  "death",      "Patient Died"
)

dictionary
```

In practice, you would load this from a CSV or define it once at the top of your analysis script.

## Labeling gtsummary Tables

### Pass the Dictionary Explicitly

```{r}
trial |>
  tbl_summary(by = trt, include = c(age, grade, marker)) |>
  add_auto_labels(dictionary = dictionary) |>
  extras()
```

### Automatic Discovery

If a `dictionary` object exists in your environment, `add_auto_labels()` finds it without you passing it:

```{r}
# dictionary already exists from above
trial |>
  tbl_summary(by = trt, include = c(age, stage, response)) |>
  add_auto_labels() |>
  extras()
```

### Pre-Labeled Data

If your data already has label attributes (e.g., from `haven::read_sas()` or manual assignment), `add_auto_labels()` reads those directly:

```{r}
labeled_trial <- trial
attr(labeled_trial$age, "label") <- "Patient Age at Baseline"
attr(labeled_trial$marker, "label") <- "Biomarker Concentration (ng/mL)"

labeled_trial |>
  tbl_summary(by = trt, include = c(age, marker)) |>
  add_auto_labels()
```

### Manual Overrides Always Win

Labels set via `label = list(...)` in `tbl_summary()` always take priority over dictionary or attribute labels:

```{r}
trial |>
  tbl_summary(
    by = trt,
    include = c(age, grade, marker),
    label = list(age ~ "Age (Custom Label)")
  ) |>
  add_auto_labels(dictionary = dictionary) |>
  extras()
```

### Regression Tables

Works with `tbl_regression()` the same way:

```{r}
lm(marker ~ age + grade + stage, data = trial) |>
  tbl_regression() |>
  add_auto_labels(dictionary = dictionary)
```

## Label Priority

When both dictionary labels and attribute labels exist for the same variable, attribute labels always take priority:
1. **Manual labels** (from `label = list(...)` in `tbl_summary()`) always win
2. **Attribute labels** (from `attr(data$var, "label")`) take priority over dictionary
3. **Dictionary labels** are used as a fallback

```{r}
trial_both <- trial
attr(trial_both$age, "label") <- "Age from Attribute"

dictionary_conflict <- tibble::tribble(
  ~Variable, ~Description,
  "age", "Age from Dictionary"
)

# Attribute wins over dictionary
trial_both |>
  tbl_summary(by = trt, include = age) |>
  add_auto_labels(dictionary = dictionary_conflict) |>
  extras()
```

## Automatic Labeling via Options

If you always keep a `dictionary` in your environment, you can skip calling `add_auto_labels()` entirely. Set this once per session (or put it in your `.Rprofile`):

```{r, eval=FALSE}
options(sumExtras.auto_labels = TRUE)
```

Now every `extras()` call picks up the dictionary automatically:

```{r, eval=FALSE}
dictionary <- tibble::tribble(
  ~Variable,    ~Description,
  "age",        "Age at Enrollment (years)",
  "marker",     "Marker Level (ng/mL)",
  "grade",      "Tumor Grade"
)

# No add_auto_labels() needed
trial |>
  tbl_summary(by = trt) |>
  extras()
```

If no dictionary is found and the data has no label attributes, `extras()` continues normally. If something goes wrong, it warns and moves on. You can still call `add_auto_labels()` explicitly whenever you need per-table control.

See `vignette("options")` for more on `.Rprofile` setup.

## More Vignettes

- `vignette("sumExtras-intro")` -- getting started with extras()
- `vignette("styling")` -- group headers and advanced formatting
- `vignette("themes")` -- JAMA compact themes
