---
title: "Package Options"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Package Options}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Auto-Labeling with `sumExtras.auto_labels`

If you keep a `dictionary` object in your environment (or your data already has label attributes), you can skip calling `add_auto_labels()` on every table. Set this once:

```{r, eval=FALSE}
options(sumExtras.auto_labels = TRUE)
```

Now `extras()` handles labeling automatically:

```{r, eval=FALSE}
# Define your dictionary once
dictionary <- tibble::tribble(
  ~Variable,    ~Description,
  "age",        "Age at Enrollment (years)",
  "marker",     "Marker Level (ng/mL)",
  "grade",      "Tumor Grade"
)

# Every extras() call picks it up
trial |>
  tbl_summary(by = trt) |>
  extras()
```

Put `options(sumExtras.auto_labels = TRUE)` in your `.Rprofile` to enable this for every session. If no dictionary is found and the data has no label attributes, `extras()` continues normally. If something goes wrong, it warns and moves on.

You can still call `add_auto_labels()` explicitly whenever you need per-table control.

## More Vignettes

- `vignette("sumExtras-intro")` -- getting started with extras()
- `vignette("labeling")` -- dictionary-based labeling and cross-package workflows
- `vignette("themes")` -- JAMA compact themes for gtsummary and gt tables
- `vignette("styling")` -- group headers and advanced formatting
