---
title: "Introduction to sumExtras"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to sumExtras}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# Set gtsummary print engine for proper rendering
options(gtsummary.print_engine = "gt")
```

```{r setup}
library(sumExtras)
library(gtsummary)
library(dplyr)

# Apply the recommended JAMA theme
use_jama_theme()
```

## Overview

If you've worked with gtsummary before, you're familiar with the typical workflow of building summary tables: creating a base table with `tbl_summary()`, then progressively adding features like overall columns, p-values, and formatting tweaks. While gtsummary's modular approach provides flexibility, the same sequence of functions appears repeatedly in analysis scripts.

sumExtras streamlines this process by providing convenience functions that apply commonly-used formatting patterns in a single step. The package handles three main pain points:

1. **Repetitive styling workflows** - Combining multiple formatting steps into one function call
2. **Inconsistent missing value displays** - Standardizing how NA values appear across tables
3. **Manual variable labeling** - Automating label assignment from data dictionaries

## The `extras()` Function

The signature function of this package, `extras()`, consolidates the most common table enhancements into a single step. At minimum, it adds bold labels, removes the "Characteristic" header, and standardizes missing value display. With default settings, it also adds an overall column and p-values.

### Basic Usage

```{r}
# Standard gtsummary workflow
trial |>
  tbl_summary(by = trt) |>
  add_overall() |>
  add_p() |>
  bold_labels() |>
  modify_header(label ~ "")

# Equivalent using extras()
trial |>
  tbl_summary(by = trt) |>
  extras()
```

Both approaches produce the same result, but `extras()` requires less code and ensures consistency across your analysis.

### Customizing Output

You can control which features are applied using the function arguments:

```{r}
# Table without p-values
trial |>
  tbl_summary(by = trt) |>
  extras(pval = FALSE)

# Table without overall column
trial |>
  tbl_summary(by = trt) |>
  extras(overall = FALSE)

# Overall column as last column (default is to set it as first)
trial |>
  tbl_summary(by = trt) |>
  extras(last = TRUE)
```

For projects with consistent table formatting requirements, you can define styling parameters once and reuse them:

```{r}
# Define standard table settings for a project
standard_table_args <- list(
  pval = TRUE,
  overall = TRUE,
  last = TRUE
)

# Apply consistently across multiple tables
trial |>
  select(age, grade, stage, trt) |>
  tbl_summary(by = trt) |>
  extras(.args = standard_table_args)
```

## Cleaning Missing Values

One subtle but important aspect of table presentation is how missing or undefined values are displayed. gtsummary tables can show various representations of missing data: "0 (NA%)", "NA (NA)", "NA, NA", etc. These inconsistencies create visual clutter and make tables harder to scan.

The `clean_table()` function (which is called automatically by `extras()`) standardizes all zero (`0 (0%)`) or missing value representations to "---":

```{r}
# Create data with some missing patterns
trial_missing <- trial |>
  mutate(
    age = if_else(trt == 'Drug B', NA_real_, age),
    marker = if_else(trt == 'Drug A', NA_real_, marker)
  )

# Without cleaning
trial_missing |>
  tbl_summary(by = trt)

# With clean_table()
trial_missing |>
  tbl_summary(by = trt) |>
  clean_table()
```

You can also use `clean_table()` independently if you prefer to build tables step-by-step:

```{r}
#| warning: false
#| message: false
trial_missing |>
  tbl_summary(by = trt) |>
  add_overall() |>
  add_p() |>
  clean_table()
```

## Automatic Variable Labeling

One of the most time-consuming aspects of creating publication-ready tables is labeling variables with human-readable descriptions. Instead of manually typing labels for every variable in every table, sumExtras provides a streamlined two-function labeling system:

- **`add_auto_labels()`** - Smart labeling for gtsummary tables
- **`apply_labels_from_dictionary()`** - Set labels as data attributes for cross-package workflows

Both functions use the same dictionary format, making it easy to maintain consistent labels across tables and plots throughout your analysis.

### Creating a Data Dictionary

A data dictionary serves dual purposes: it documents your variables and provides labels for automatic application. The dictionary is simply a data frame with two required columns:

- **`Variable`**: The exact variable names from your dataset
- **`Description`**: Human-readable labels you want to display

```{r}
# Create a dictionary for the trial dataset
dictionary <- tibble::tribble(
  ~Variable,    ~Description,
  "trt",        "Chemotherapy Treatment",
  "age",        "Age at Enrollment (years)",
  "marker",     "Marker Level (ng/mL)",
  "stage",      "T Stage",
  "grade",      "Tumor Grade",
  "response",   "Tumor Response"
)
```

In real projects, you would typically maintain this dictionary as a CSV file or database table, loading it once at the beginning of your analysis. This centralizes your variable documentation and ensures consistency across all outputs.

### Labeling gtsummary Tables with `add_auto_labels()`

The `add_auto_labels()` function is designed to be flexible and intelligent. It can work with dictionaries, pre-labeled data, or both, and it always respects manual overrides.

#### Basic Usage: Pass Dictionary Explicitly

The most straightforward approach is to pass your dictionary directly to the function:

```{r}
trial |>
  tbl_summary(by = trt, include = c(age, grade, marker)) |>
  add_auto_labels(dictionary = dictionary) |>
  extras()
```

#### Automatic Discovery: Dictionary in Environment

If you have a `dictionary` object in your environment, `add_auto_labels()` will find it automatically without needing to pass it explicitly:

```{r}
# Dictionary is already in environment from above
trial |>
  tbl_summary(by = trt, include = c(age, stage, response)) |>
  add_auto_labels() |>  # Finds dictionary automatically
  extras()
```

This is particularly convenient when working in an R Markdown or Quarto document where your dictionary is defined once at the top.

#### Working with Pre-Labeled Data (Attributes)

If your data already has label attributes (from packages like haven, labelled, or set manually), `add_auto_labels()` can read them directly:

```{r}
# Create data with label attributes
labeled_trial <- trial
attr(labeled_trial$age, "label") <- "Patient Age at Baseline"
attr(labeled_trial$marker, "label") <- "Biomarker Concentration (ng/mL)"

# Use attributes for labeling (no dictionary needed)
labeled_trial |>
  tbl_summary(by = trt, include = c(age, marker)) |>
  add_auto_labels()  # Reads from label attributes
```

#### Manual Overrides Always Win

No matter where labels come from (dictionary or attributes), manual labels specified in your `tbl_summary()` call always take precedence:

```{r}
trial |>
  tbl_summary(
    by = trt,
    include = c(age, grade, marker),
    label = list(age ~ "Age (Custom Label)")  # This overrides dictionary/attributes
  ) |>
  add_auto_labels(dictionary = dictionary) |>
  extras()
```

This gives you complete control: use automated labeling for most variables, but override specific ones when needed.

#### Working with Regression Tables

The labeling system works seamlessly with regression tables too:

```{r}
lm(marker ~ age + grade + stage, data = trial) |>
  tbl_regression() |>
  add_auto_labels(dictionary = dictionary)
```

### Controlling Label Priority with Options

When your data has both dictionary labels and attribute labels available, `add_auto_labels()` needs to decide which one to use. You control this with a global option.

#### Default Behavior: Attributes Have Priority

By default, label attributes take precedence over dictionary labels. This respects labels that may have been carefully set by data import functions (like `haven::read_sas()`) or other preprocessing steps:

```{r}
# Create data with both sources of labels
trial_both <- trial
attr(trial_both$age, "label") <- "Age from Attribute"

# Also have dictionary (already defined above)
dictionary_conflict <- tibble::tribble(
  ~Variable, ~Description,
  "age", "Age from Dictionary"
)

# Default: attribute wins
trial_both |>
  tbl_summary(by = trt, include = age) |>
  add_auto_labels(dictionary = dictionary_conflict) |>
  extras()
# Shows: "Age from Attribute"
```

#### Prefer Dictionary: When to Use `TRUE`

If you want dictionary labels to override attribute labels, set the `sumExtras.preferDictionary` option to `TRUE`. This is useful when you're actively maintaining a master dictionary and want it to be the single source of truth:

```{r}
# Prioritize dictionary over attributes
options(sumExtras.preferDictionary = TRUE)

trial_both |>
  tbl_summary(by = trt, include = age) |>
  add_auto_labels(dictionary = dictionary_conflict) |>
  extras()
# Shows: "Age from Dictionary"

# Reset to default for rest of vignette
options(sumExtras.preferDictionary = FALSE)
```

**When to use each setting:**

- **`FALSE` (default)**: You're importing labeled data from SAS/Stata/SPSS and want to preserve those labels
- **`TRUE`**: You maintain a master dictionary and want it to override any existing labels

Remember: manual labels set via `label = list(...)` always win, regardless of this option.

### Cross-Package Workflows: One Dictionary for Tables and Plots

Often you need consistent labels across both gtsummary tables and ggplot2 visualizations. The `apply_labels_from_dictionary()` function enables this by setting dictionary labels as data attributes, which ggplot2 4.0+ can use automatically.

#### Setting Labels Once, Using Everywhere

Apply your dictionary labels to the data once, then use the labeled data for both tables and plots:

```{r}
# Apply labels to data as attributes
trial_labeled <- trial |>
  apply_labels_from_dictionary(dictionary = dictionary)

# Now use for gtsummary table
trial_labeled |>
  tbl_summary(by = trt, include = c(age, marker, grade)) |>
  add_auto_labels() |>  # Reads attributes automatically
  extras()
```

#### ggplot2 4.0+ Integration

With ggplot2 version 4.0 and later, label attributes are automatically used for axis and legend labels:

```{r, fig.width=7, fig.height=4}
library(ggplot2)

# Labels appear automatically on axes and legend!
trial_labeled |>
  ggplot(aes(x = age, y = marker, color = trt)) +
  geom_point(alpha = 0.6) +
  theme_minimal()
```

No need to manually specify `labs()` - the labels from your dictionary are applied automatically.

#### Complete Workflow: Dictionary to Tables and Plots

Here's a realistic workflow showing how one dictionary serves both gtsummary tables and ggplot2 visualizations:

```{r, fig.width=7, fig.height=5}
# 1. Define dictionary once
my_dictionary <- tibble::tribble(
  ~Variable,    ~Description,
  "age",        "Age at Enrollment (years)",
  "marker",     "Marker Level (ng/mL)",
  "trt",        "Treatment Group",
  "grade",      "Tumor Grade",
  "stage",      "T Stage"
)

# 2. Apply to data
trial_final <- trial |>
  apply_labels_from_dictionary(my_dictionary)

# 3. Create gtsummary table
trial_final |>
  tbl_summary(
    by = trt,
    include = c(age, marker, grade, stage)
  ) |>
  add_auto_labels() |>
  extras()

# 4. Create ggplot2 visualization with same labels
trial_final |>
  filter(!is.na(marker)) |>
  ggplot(aes(x = age, y = marker)) +
  geom_point(aes(color = grade), alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  facet_wrap(~trt) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Notice how the axis labels and legend titles are automatically pulled from your dictionary - no manual `labs()` calls needed!

### Real-World Example: Complete Analysis

Here's a comprehensive example showing how the labeling system streamlines a typical analysis workflow:

```{r, fig.width=8, fig.height=6}
# Step 1: Define your master dictionary
# In practice, this would be loaded from a CSV file
study_dictionary <- tibble::tribble(
  ~Variable,    ~Description,
  "trt",        "Treatment Assignment",
  "age",        "Age at Baseline (years)",
  "marker",     "Biomarker Level (ng/mL)",
  "stage",      "Clinical Stage",
  "grade",      "Tumor Grade",
  "response",   "Treatment Response"
)

# Step 2: Apply labels to your data once
trial_study <- trial |>
  apply_labels_from_dictionary(study_dictionary)

# Step 3: Create multiple tables using the same labels
# Table 1: Overall summary
trial_study |>
  tbl_summary(include = c(age, marker, stage, grade)) |>
  add_auto_labels() |>
  extras(overall = TRUE, pval = FALSE)

# Table 2: By treatment comparison
trial_study |>
  tbl_summary(
    by = trt,
    include = c(age, marker, response)
  ) |>
  add_auto_labels() |>
  extras()

# Step 4: Create plots using the same labels
# Plot 1: Age distribution by treatment
trial_study |>
  ggplot(aes(x = trt, y = age, fill = trt)) +
  geom_boxplot(alpha = 0.7) +
  theme_minimal() +
  theme(legend.position = "none")

# Plot 2: Marker vs age relationship
trial_study |>
  filter(!is.na(marker)) |>
  ggplot(aes(x = age, y = marker, color = trt)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = FALSE) +
  theme_minimal()
```

This workflow demonstrates the power of the labeling system: define your labels once in a dictionary, apply them to your data, then create as many tables and plots as you need with consistent, professional labeling throughout.

## Styling Group Headers

When you organize variables into sections using `add_variable_group_header()`, the `group_styling()` function adds visual emphasis to make those section headers stand out:

```{r}
trial |>
  select(age, marker, grade, stage, response, trt) |>
  tbl_summary(by = trt) |>
  extras() |>
  add_variable_group_header(
    header = "Demographics",
    variables = age
  ) |>
  add_variable_group_header(
    header = "Clinical Measures",
    variables = marker:response
  ) |>
  group_styling()
```

By default, this applies both bold and italic formatting to group headers. You can customize the text formatting:

```{r}
trial |>
  select(age, marker, grade, stage, trt) |>
  tbl_summary(by = trt) |>
  extras() |>
  add_variable_group_header(
    header = "Patient Characteristics",
    variables = age:stage
  ) |>
  group_styling(format = "bold")  # Bold only
```

Using `group_styling()` also restores the original left-justified variable label indentation. The default behavior of `add_variable_group_header()` is to align the variable label with categorical levels or "Unknown" display.

### Adding Background Color to Group Headers

For additional visual distinction, you can add a gray background to group headers after converting to a gt table. Use the `get_group_rows()` helper function to identify which rows contain group headers:

```{r}
# Create table with groups and styling
my_table <- trial |>
  select(age, marker, grade, stage, response, trt) |>
  tbl_summary(by = trt) |>
  extras() |>
  add_variable_group_header(
    header = "Demographics",
    variables = age
  ) |>
  add_variable_group_header(
    header = "Clinical Measures",
    variables = marker:response
  ) |>
  group_styling()

# Get group row numbers before converting to gt
group_rows <- get_group_rows(my_table)

# Convert to gt and apply gray background
my_table |>
  as_gt() |>
  gt::tab_style(
    style = gt::cell_fill(color = "#E8E8E8"),
    locations = gt::cells_body(rows = group_rows)
  )
```

This pattern combines text formatting (bold/italic) with background color (#E8E8E8, a light gray) to create clear visual separation between table sections.

## Theme Management

### Applying the JAMA Theme

sumExtras is designed to work best with the JAMA compact theme. Use `use_jama_theme()` to apply this theme to all gtsummary tables in your session:

```{r}
# Apply JAMA compact theme (typically done once at the beginning)
use_jama_theme()
```

This is equivalent to calling `gtsummary::set_gtsummary_theme(gtsummary::theme_gtsummary_compact("jama"))` but provides a more convenient interface. You can reset to the default gtsummary theme at any time with `gtsummary::reset_gtsummary_theme()`.

### Compact Themes for gt Tables

When mixing gtsummary tables with regular gt tables in the same document, visual consistency matters. The `theme_gt_compact()` function applies JAMA-style compact formatting to gt tables, matching the appearance of the JAMA compact theme:

```{r}
# gtsummary table (uses the theme set with use_jama_theme())
trial |>
  select(trt, age, grade) |>
  tbl_summary() |>
  extras()

# Regular gt table with matching compact style
trial |>
  select(trt, age, grade) |>
  head(10) |>
  gt::gt() |>
  theme_gt_compact()
```

## Putting It All Together

Here's a complete workflow demonstrating how these functions work together:

```{r}
# 1. Define your dictionary (typically done once per project)
dictionary <- tibble::tribble(
  ~Variable,    ~Description,
  "trt",        "Chemotherapy Treatment",
  "age",        "Age at Enrollment (years)",
  "marker",     "Marker Level (ng/mL)",
  "stage",      "T Stage",
  "grade",      "Tumor Grade",
  "response",   "Tumor Response",
  "death",      "Patient Died"
)

# 2. Set the recommended theme
use_jama_theme()

# 3. Create a table with automatic labels and standard formatting
trial |>
  select(trt, age, marker, stage, grade, response) |>
  tbl_summary(
    by = trt,
    missing = "no"
  ) |>
  add_auto_labels() |>
  extras()

# 4. Table with grouped variables, styling, and gray background
final_table <- trial |>
  select(trt, age, marker, grade, stage, response) |>
  tbl_summary(by = trt, missing = "no") |>
  add_auto_labels() |>
  extras() |>
  add_variable_group_header(
    header = "Baseline Characteristics",
    variables = age:marker
  ) |>
  add_variable_group_header(
    header = "Disease Characteristics",
    variables = grade:response
  ) |>
  group_styling()

# Get group rows and apply gray background
group_rows <- get_group_rows(final_table)

final_table |>
  as_gt() |>
  gt::tab_style(
    style = gt::cell_fill(color = "#E8E8E8"),
    locations = gt::cells_body(rows = group_rows)
  )
```

## Next Steps

This vignette covered the core functionality of sumExtras. For detailed information about individual functions, see their help documentation:

- `?extras` - Main styling function
- `?clean_table` - Missing value standardization
- `?add_auto_labels` - Automatic variable labeling for gtsummary tables
- `?apply_labels_from_dictionary` - Set label attributes for cross-package workflows
- `?use_jama_theme` - Apply JAMA compact theme
- `?group_styling` - Group header formatting
- `?get_group_rows` - Get row numbers of variable group headers
- `?theme_gt_compact` - Compact gt table themes

The package is designed to reduce repetitive code while maintaining the flexibility of gtsummary's modular approach. Use as much or as little as fits your workflow.
