---
title: "Introduction to sumExtras"
format: html
vignette: >
  %\VignetteIndexEntry{Introduction to sumExtras}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(sumExtras)
library(gtsummary)
library(dplyr)

# Apply the recommended JAMA theme
use_jama_theme()
```

## Overview

If you've worked with gtsummary before, you're familiar with the typical workflow of building summary tables: creating a base table with `tbl_summary()`, then progressively adding features like overall columns, p-values, and formatting tweaks. While gtsummary's modular approach provides flexibility, the same sequence of functions appears repeatedly in analysis scripts.

sumExtras streamlines this process by providing convenience functions that apply commonly-used formatting patterns in a single step. The package handles three main pain points:

1. **Repetitive styling workflows** - Combining multiple formatting steps into one function call
2. **Inconsistent missing value displays** - Standardizing how NA values appear across tables
3. **Manual variable labeling** - Automating label assignment from data dictionaries

## The `extras()` Function

The signature function of this package, `extras()`, consolidates the most common table enhancements into a single step. At minimum, it adds bold labels, removes the "Characteristic" header, and standardizes missing value display. With default settings, it also adds an overall column and p-values.

### Basic Usage

```{r}
# Standard gtsummary workflow
trial |>
  tbl_summary(by = trt) |>
  add_overall() |>
  add_p() |>
  bold_labels() |>
  modify_header(label ~ "")

# Equivalent using extras()
trial |>
  tbl_summary(by = trt) |>
  extras()
```

Both approaches produce the same result, but `extras()` requires less code and ensures consistency across your analysis.

### Customizing Output

You can control which features are applied using the function arguments:

```{r}
# Table without p-values
trial |>
  tbl_summary(by = trt) |>
  extras(pval = FALSE)

# Table without overall column
trial |>
  tbl_summary(by = trt) |>
  extras(overall = FALSE)

# Overall column as last column (default is to set it as first)
trial |>
  tbl_summary(by = trt) |>
  extras(last = TRUE)
```

For projects with consistent table formatting requirements, you can define styling parameters once and reuse them:

```{r}
# Define standard table settings for a project
standard_table_args <- list(
  pval = TRUE,
  overall = TRUE,
  last = TRUE
)

# Apply consistently across multiple tables
trial |>
  select(age, grade, stage, trt) |>
  tbl_summary(by = trt) |>
  extras(.args = standard_table_args)
```

## Cleaning Missing Values

One subtle but important aspect of table presentation is how missing or undefined values are displayed. gtsummary tables can show various representations of missing data: "0 (NA%)", "NA (NA)", "NA, NA", etc. These inconsistencies create visual clutter and make tables harder to scan.

The `clean_table()` function (which is called automatically by `extras()`) standardizes all zero (`0 (0%)`) or missing value representations to "---":

```{r}
# Create data with some missing patterns
trial_missing <- trial |>
  mutate(
    age = if_else(trt == 'Drug B', NA_real_, age),
    marker = if_else(trt == 'Drug A', NA_real_, marker)
  )

# Without cleaning
trial_missing |>
  tbl_summary(by = trt)

# With clean_table()
trial_missing |>
  tbl_summary(by = trt) |>
  clean_table()
```

You can also use `clean_table()` independently if you prefer to build tables step-by-step:

```{r}
#| warning: false
#| message: false
trial_missing |>
  tbl_summary(by = trt) |>
  add_overall() |>
  add_p() |>
  clean_table()
```

## Automatic Variable Labeling

For projects with many variables, manually specifying labels for each one becomes tedious and error-prone. The `add_auto_labels()` function automates this process by pulling labels from a data dictionary.

### Setting Up a Dictionary

The function expects a dictionary object in your global environment with two columns: `Variable` (exact variable names) and `Description` (human-readable labels):

```{r}
# Create a dictionary for the trial dataset
dictionary <- tibble::tribble(
  ~Variable,    ~Description,
  "trt",        "Chemotherapy Treatment",
  "age",        "Age at Enrollment",
  "marker",     "Marker Level (ng/mL)",
  "stage",      "T Stage",
  "grade",      "Tumor Grade",
  "response",   "Tumor Response"
)
```

In practice, you would typically maintain this dictionary as a CSV file or within a data management script, loading it once at the beginning of your analysis.

### Applying Automatic Labels

Once your dictionary is defined, pipe any gtsummary table through `add_auto_labels()`:

```{r}
trial |>
  tbl_summary(by = trt) |>
  add_auto_labels() |>
  extras()
```

The function intelligently handles several scenarios:

- Variables not in the dictionary are left unlabeled (showing the variable name)
- Variables not included in the table are ignored
- Manual label overrides specified in `tbl_summary(label = ...)` take precedence over dictionary labels

```{r}
# Manual override takes precedence
trial |>
  tbl_summary(
    by = trt,
    label = list(age ~ "Patient Age (years)")  # This overrides dictionary
  ) |>
  add_auto_labels() |>
  extras()
```

### Working with Regression Tables

The labeling system also works with regression tables:

```{r}
lm(marker ~ age + grade + stage, data = trial) |>
  tbl_regression() |>
  add_auto_labels()
```

## Styling Group Headers

When you organize variables into sections using `add_variable_group_header()`, the `group_styling()` function adds visual emphasis to make those section headers stand out:

```{r}
trial |>
  select(age, marker, grade, stage, response, trt) |>
  tbl_summary(by = trt) |>
  extras() |>
  add_variable_group_header(
    header = "Demographics",
    variables = age
  ) |>
  add_variable_group_header(
    header = "Clinical Measures",
    variables = marker:response
  ) |>
  group_styling()
```

By default, this applies both bold and italic formatting to group headers. You can customize the text formatting:

```{r}
trial |>
  select(age, marker, grade, stage, trt) |>
  tbl_summary(by = trt) |>
  extras() |>
  add_variable_group_header(
    header = "Patient Characteristics",
    variables = age:stage
  ) |>
  group_styling(format = "bold")  # Bold only
```

### Adding Background Color to Group Headers

For additional visual distinction, you can add a gray background to group headers after converting to a gt table. Use the `get_group_rows()` helper function to identify which rows contain group headers:

```{r}
# Create table with groups and styling
my_table <- trial |>
  select(age, marker, grade, stage, response, trt) |>
  tbl_summary(by = trt) |>
  extras() |>
  add_variable_group_header(
    header = "Demographics",
    variables = age
  ) |>
  add_variable_group_header(
    header = "Clinical Measures",
    variables = marker:response
  ) |>
  group_styling()

# Get group row numbers before converting to gt
group_rows <- get_group_rows(my_table)

# Convert to gt and apply gray background
my_table |>
  as_gt() |>
  gt::tab_style(
    style = gt::cell_fill(color = "#E8E8E8"),
    locations = gt::cells_body(rows = group_rows)
  )
```

This pattern combines text formatting (bold/italic) with background color (#E8E8E8, a light gray) to create clear visual separation between table sections.

## Theme Management

### Applying the JAMA Theme

sumExtras is designed to work best with the JAMA compact theme. Use `use_jama_theme()` to apply this theme to all gtsummary tables in your session:

```{r}
# Apply JAMA compact theme (typically done once at the beginning)
use_jama_theme()
```

This is equivalent to calling `gtsummary::set_gtsummary_theme(gtsummary::theme_gtsummary_compact("jama"))` but provides a more convenient interface. You can reset to the default gtsummary theme at any time with `gtsummary::reset_gtsummary_theme()`.

### Compact Themes for gt Tables

When mixing gtsummary tables with regular gt tables in the same document, visual consistency matters. The `theme_gt_compact()` function applies JAMA-style compact formatting to gt tables, matching the appearance of the JAMA compact theme:

```{r}
# gtsummary table (uses the theme set with use_jama_theme())
summary_table <- trial |>
  select(trt, age, grade) |>
  tbl_summary() |>
  extras()

# Regular gt table with matching compact style
trial |>
  select(trt, age, grade) |>
  head(10) |>
  gt::gt() |>
  theme_gt_compact()
```

## Putting It All Together

Here's a complete workflow demonstrating how these functions work together:

```{r}
# 1. Define your dictionary (typically done once per project)
dictionary <- tibble::tribble(
  ~Variable,    ~Description,
  "trt",        "Chemotherapy Treatment",
  "age",        "Age at Enrollment (years)",
  "marker",     "Marker Level (ng/mL)",
  "stage",      "T Stage",
  "grade",      "Tumor Grade",
  "response",   "Tumor Response",
  "death",      "Patient Died"
)

# 2. Set the recommended theme
use_jama_theme()

# 3. Create a table with automatic labels and standard formatting
trial |>
  select(trt, age, marker, stage, grade, response) |>
  tbl_summary(
    by = trt,
    missing = "no"
  ) |>
  add_auto_labels() |>
  extras()

# 4. Table with grouped variables, styling, and gray background
final_table <- trial |>
  select(trt, age, marker, grade, stage, response) |>
  tbl_summary(by = trt, missing = "no") |>
  add_auto_labels() |>
  extras() |>
  add_variable_group_header(
    header = "Baseline Characteristics",
    variables = age:marker
  ) |>
  add_variable_group_header(
    header = "Disease Characteristics",
    variables = grade:response
  ) |>
  group_styling()

# Get group rows and apply gray background
group_rows <- get_group_rows(final_table)

final_table |>
  as_gt() |>
  gt::tab_style(
    style = gt::cell_fill(color = "#E8E8E8"),
    locations = gt::cells_body(rows = group_rows)
  )
```

## Next Steps

This vignette covered the core functionality of sumExtras. For detailed information about individual functions, see their help documentation:

- `?extras` - Main styling function
- `?clean_table` - Missing value standardization
- `?add_auto_labels` - Automatic variable labeling
- `?create_labels` - Manual label list creation
- `?use_jama_theme` - Apply JAMA compact theme
- `?group_styling` - Group header formatting
- `?get_group_rows` - Get row numbers of variable group headers
- `?theme_gt_compact` - Compact gt table themes

The package is designed to reduce repetitive code while maintaining the flexibility of gtsummary's modular approach. Use as much or as little as fits your workflow.
