---
title: "Themes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Themes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options(gtsummary.print_engine = "gt")
```

```{r setup}
#| eval: false
library(sumExtras)
library(gtsummary)
library(gt)
```

```{r setup2}
#| echo: false
#| message: false
#| warning: false
library(sumExtras)
library(gtsummary)
library(gt)
```

## `use_jama_theme()`

`use_jama_theme()` sets the gtsummary theme to JAMA compact styling for the rest of your session. It reduces padding, tightens font sizes, and produces tables suited for publication or reports.

```{r}
use_jama_theme()

trial |>
  tbl_summary(by = trt) |>
  extras()
```

This is equivalent to calling `gtsummary::set_gtsummary_theme(gtsummary::theme_gtsummary_compact("jama"))` but shorter to type.

To reset back to the default gtsummary theme:

```{r, eval=FALSE}
gtsummary::reset_gtsummary_theme()
```

## `theme_gt_compact()`

When you mix gtsummary tables with plain gt tables in the same document, the styling mismatch is noticeable. `theme_gt_compact()` applies the same JAMA compact look to gt tables.

:::::: {style="display: flex; gap: 15px; margin-bottom: 20px; align-items: flex-start;"}

::: {style="flex: 1; max-width: 48%;"}

#### Default gt

```{r gt-default, eval=FALSE}
trial |>
  select(trt, age, grade) |>
  head(10) |>
  gt::gt()
```

:::

::: {style="flex: 1; max-width: 48%;"}

#### With theme_gt_compact()

```{r gt-compact, eval=FALSE}
trial |>
  select(trt, age, grade) |>
  head(10) |>
  gt::gt() |>
  theme_gt_compact()
```

:::

::::::

```{r build-gt-theme, echo=FALSE}
table_gt_default <- trial |>
  dplyr::select(trt, age, grade) |>
  head(10) |>
  gt::gt()

table_gt_compact <- trial |>
  dplyr::select(trt, age, grade) |>
  head(10) |>
  gt::gt() |>
  theme_gt_compact()
```

:::::: {style="display: flex; gap: 15px; margin-bottom: 30px; align-items: flex-start;"}

::: {style="flex: 1; max-width: 48%;"}

```{r render-gt-default, echo=FALSE}
table_gt_default
```

:::

::: {style="flex: 1; max-width: 48%;"}

```{r render-gt-compact, echo=FALSE}
table_gt_compact
```

:::

::::::

You can layer additional gt styling on top:

```{r}
trial |>
  dplyr::select(trt, age, grade, marker) |>
  head(8) |>
  gt::gt() |>
  theme_gt_compact() |>
  gt::tab_header(
    title = "Trial Patient Sample",
    subtitle = "First 8 patients"
  )
```

## More Vignettes

- `vignette("sumExtras-intro")` -- getting started with extras()
- `vignette("labeling")` -- dictionary-based labeling
- `vignette("styling")` -- group headers and advanced formatting
- `vignette("options")` -- .Rprofile options for automatic labeling
