---
title: "sumExtras: Streamlining Table Workflows in Clinical Research with R"
author:
  - name: "Kyle Grealis"
    affiliation: "University of Miami"
date: "May 8, 2026"
date-format: "MMM D, YYYY"
---

```{r}
#| echo: false
suppressPackageStartupMessages({
  library(tidyverse)
  library(sumExtras)
  library(gtsummary)
  library(gt)
})
```

# Introduction
 
> **sumExtras**: "gt**sum**mary table **Extras**"


## The plan
::: {.incremental}
+ Apply a JAMA-ready theme
+ Customize missing values with `clean_table()`
+ One-line table formatting with `extras()`
+ Auto-label tables with `add_auto_labels()`
  + Create a variable dictionary
  + Set package options
+ Add variable group headers
+ Style and color group rows
+ Put it all together in one pipeline
:::


## What we'll build

(insert final table image here)


## Creating a summary table

```{r}
#| echo: true
#| eval: true
trial
```


## Creating a Variable Dictionary

A dictionary object is a data frame with two columns: `variable` and `description`.

<!-- Show the creation of the data dictionary object
Highlight lines, then show the created object incrementally to the side -->

:::: {.columns}
::: {.column width="55%"}
```{r}
#| echo: true
#| code-line-numbers: "|1|2"
dictionary <- tibble::tribble(
  ~variable,    ~description,
  "trt",        "Chemotherapy Treatment",
  "age",        "Age at Enrollment (years)",
  "marker",     "Marker Level (ng/mL)",
  "stage",      "T Stage",
  "grade",      "Tumor Grade",
  "response",   "Tumor Response",
  "death",      "Patient Died"
)
```
:::
::: {.column width="45%" .fragment}
```{r}
dictionary
```
:::
::::

::: {.incremental}
::: {.callout-note .fragment}
The column names are case-*in*sensitive. `variable`, `Variable`, and `VARIABLE`
will work all the same.
:::
:::


<!-- SLIDE OUTLINE
1. INTRO: Show verbose gtsummary-only code (the "before")
   - tbl_summary(by=trt) |> add_overall() |> add_p() |> bold_labels() |>
     bold_p() |> modify_header(...) |> modify_table_body(...)
   - Preview the final sumExtras version we're building toward

2. THEME: use_jama_theme() before/after comparison
   - Show default gtsummary output vs JAMA-themed output

3. CLEAN TABLE: clean_table() for missing values
   - Show table with ugly NA/0 patterns, then pipe on clean_table()

4. EXTRAS: extras() replaces repeated boilerplate
   - Show how one call replaces add_overall, add_p, bold_labels, bold_p,
     modify_header, clean_table
   - Mention key arguments: pval, overall, last, header, symbol

5. AUTO LABELS: add_auto_labels() with dictionary
   - Show dictionary creation (already built above)
   - Demonstrate label priority: manual > attributes > dictionary
   - Show package options: sumExtras.auto_labels, sumExtras.prefer_dictionary

6. GROUP HEADERS: add_variable_group_header()
   - Add grouping to organize variables in the table

7. GROUP STYLING & COLORS: add_group_styling() + add_group_colors()
   - Style group rows, add color — visual payoff
   - add_group_colors() must be last (converts to gt)

8. OUTRO: Full pipeline side-by-side
   - Verbose gtsummary code vs clean sumExtras pipeline
   - Circle back to the "before" from slide 1
-->
